Why does female meiosis allow the opportunity for exploitation by selfish sperm? BETTER TITLE?

### 2013-09-11 writeup_and_notes/male_dependent_drive.tex
finished a very rough draft of results, and added figs

%%!TEX TS-program = latex
\documentclass[12pt,letterpaper]{article}
\usepackage{epsfig}                 
\usepackage[authoryear]{natbib}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{psfrag}
\usepackage{mathabx}
\renewcommand{\baselinestretch}{1.6}
\large
\pagenumbering{arabic}
\usepackage[usenames,dvipsnames]{color}
\usepackage{fullpage}
\bibliographystyle{genetics}
\usepackage{multirow}
\newcommand{\ye}{\hat{y}}
\newcommand{\xe}{\hat{x}}
\usepackage{color}
\usepackage[normalem]{ulem}

\title{Sperm modifiers of female drive.}
\author{Yaniv Brandvain \\ email: ybrandvain@gmail.com  \and Graham Coop \\ email: gmcoop@ucdavis.edu }

@@

%Since all alleles within an individual rely on individual survival and reproduction for evolutionary success, most of life in a diploid eukaryotic genome is harmonious.
%However, this harmony is incomplete, as in numerous cases an allele can benefit (in the short or long term) from taking advantage of its host.
Despite the apparent unity of the organism, occasionally alleles can
gain an evolutionary advantage at a cost to  individual fitness
\citep{Burt2006}, often by exploiting meiosis and gametogensis.
%A clear opportunity for this conflict arises during gametogenesis where alternative alleles are in direct competition for representation in the next generation.
Female meiosis, an asymmetric event in which only one of two alternate alleles
enters the  egg, with the others consigned to the polar body, is one such occasion \citep{Sandler1957,Pardo-ManuelDeVillena2001a}. 
%Female meiosis is particularly unfair, only one of the four products of meiosis becomes an egg, while the other threes are disgarded into polar bodies. 
An allele that biases female meiosis in its favor (i.e. a meiotic driver), may increase in frequency even if this driver entails a pleiotropic fitness cost \citep{Prout1973}, generating a genetic conflict between the success of the driver and organismal fitness.
%Alleles that can segregate into the egg more than half the time in heterozygotes, by exploiting asymmetries, 
%can potentially increase in frequency in the population (i.e. experiencing true meiotic drive).
%If such alleles have fitness consequences they are a source of genetic conflict, 
%and can become balanced in the population if their host in homozygotes 
%out weighs their ability to spread to spread through heterozygotes. 
Meiotic drivers observed in nature (in both plants
\citep{Buckler1999,Fishman2005,Fishman2008}, and animals
\citep{Agulnik1990,Wu2005,Pardo-ManuelDeVillena2001c}) highlight this
conflict -- the selfish benefits of drive and the associated
pleiotropic fitness costs sustain a balanced polymorphism
\citep{Prout1973}, 
and often generate on ongoing evolutionary escalation of drive suppressors and enhancers \citep{Dawe1996,Fishman2008}. 
%Such balanced drive systems can cause the subsequent evolution of linked enhancers of drive
%and of supressors of drive throughout the genome. 
%Indeed, the known polymorphic female meiotic drive systems (OM, maize knobs, IN, mimulus, others?) habor
%a great diversity of enhancers and supressors. 
The threat of meiotic drive to organismal fitness is potentially so
severe that it has been hypothesized that many basic properties of meiosis and o\"{o}genesis, including the initial genome doubling in meiosis I \citep{Haig1991}, the commonness of arrested female meiosis \citep{Mira1998}, and sex differences in the recombination rate \citep{Haig2010,Brandvain2012} have evolved to disrupt meiotic drive and enforce fairness.

@@

It is also not the case that sperm is mechanistically incapable of influencing the outcome of female meiosis.
%This is not just idle speculation, as there is also direct evidence that sperm components and indeed different sperm genotypes can affect female meiosis.
%are mechanistic reasons to think that this is more than idle speculation.
Mechanistic evidence for this possibility comes from \emph{C. elegans}, where experimentally suppressing XXX leads to premature deployment of the aster (a vital component of mitotic machinery) provided by the sperm, disrupting MII meiotic segregation in the egg, leading to a triploid zygote. 
%This demonstrates that physical components derived from the sperm are are capable of affecting the outcome of female meiosis.
Additionally, genetic evidence suggests that the transmission patterns in heterozygous females can depend on sperm haplotype. 
Specifically, the two best characterized female meiotic drive systems in mouse (In and Om), both operate by distorting the second meiotic division, 
and in both systems the outcome of female meiosis depends the genotype of the fertilizing sperm \citep{Agulnik1993,Wu2005}.

@@

\yb{YB: To me its (1) male enhancement of female drive cannot maintain a stable polymorphisms, and (2) An allele in sperm can evolve to suppress its drive in females.}

Rough order of rest of text
\subsection*{A single locus model} 
The standard model of female meiotic drive is two allele model, where the
allele $1$ is transmitted $d$ fraction of the time through female
meiosis ($d > \half$), regardless of the genotype of the sperm. 
In mammals, fertilization takes place at MII, so we imagine this drive must be taking place at MII in order
for a sperm to have any influence. For drive to take place at MII there has to be an uneven number of
 crossovers between the centromere and the drive locus, such that realistically $d$ is bounded to be $<0.75$?. 
In other systems where fertilization occurs during MI, sperm could influence either drive at MI or MII, 
and drivers at MI can have a $d=1$ if they occur in tight linkage with the centromere.

@@

When conditions are suitable for such alleles to spread into the
population, they spread slowly through low frequencies as they rarely
drive as they are unlikely to be fertilized by a sperm with the right
allele. Once they become established they spread rapidly due to their
positive frequency dependent behavior (Inset in Figure \ref{Invasion_space}). 
Unlike standard meiotic drive systems 
these alleles can not become balanced in the population  by homozygote
cost, as they have already over come this cost just to enter the population.

HETERO

@@

Cost of alleles

We'll setup a model where the sperm allele affects outcome of female meiosis. This may not be biologically
realistic as there is very little sperm expression. More realistically the male genotype that produced the sperm
may be more relevant, as males could place products in their sperm that influenced the outcome of female meoisis.
However, in practice models that allow the influence of fertilizing male genotype on female meiosis 
seem to behave very similiarly to those based on the allele of the fertilizing sperm.
For simplicity we'll focus on models based on the allele of the sperm, as it's bloodly complicated enough already.

\subsection*{ A) Invasion by single self promoting allele}

show that such alleles:
\begin{enumerate}
\item can't be balanced, \\
\item and homozygous problems are tested out at low freq.  \\
\item Any heterozygous problems, leads to a bistable allele\\
\item If these alleles take off they speed through to fixation\\
\item If allele has any drive ability in absence of sperm effect that is what allows it to enter the population
and sperm effect isn't a further cause of conflict. What if anything do we mean by this?\\
\item PERHAPS HERE WE INTRODUCE A SELF-RESTRAINING ALLELE
\end{enumerate}

@@

\subsection*{B) A more biologically realistic selfish sperm system}
Three allele systems. See Figure \ref{Eggsperm_3_allele_cartoon} \\ 
Perhaps these self promoting alleles are biologically unrealistic as new alleles needs 
both ability to drive in female meiosis and to influence that drive in sperm.
Perhaps more realistic to think that a female drive system evolves, 
and then a sperm modifier appears. This will only occur if original system is trapped 
at drive-selection balance. So driver must have homozygous cost.

>>>>

%%!TEX TS-program = latex
\documentclass[12pt,letterpaper]{article}
\usepackage{epsfig}                 
\usepackage[authoryear]{natbib}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{psfrag}
\usepackage{mathabx}
\renewcommand{\baselinestretch}{1.6}
\large
\pagenumbering{arabic}
\usepackage[usenames,dvipsnames]{color}
\usepackage{fullpage}
\bibliographystyle{genetics}
\usepackage{multirow}
\newcommand{\ye}{\hat{y}}
\newcommand{\xe}{\hat{x}}
\usepackage{color}
\usepackage[normalem]{ulem}  
	\newcommand{\gc}[1]{{ \color{red} #1}}
	\newcommand{\yb}[1]{{ \color{blue} #1}}

%Why do animals allow male control of female meiosis?

%Why does female meiosis allow the opportunity for exploitation by selfish sperm?

%sperm and egg genomes both favour the peaceful resolution of female meiotic drive
%strange bed fellows

\title{Why does female meiosis allow the opportunity for exploitation
  by selfish sperm? BETTER TITLE?}
\author{Yaniv Brandvain \\ email: ybrandvain@gmail.com  \and Graham Coop \\ email: gmcoop@ucdavis.edu }

@@

%Since all alleles within an individual rely on individual survival and reproduction for evolutionary success, most of life in a diploid eukaryotic genome is harmonious.
%However, this harmony is incomplete, as in numerous cases an allele can benefit (in the short or long term) from taking advantage of its host.
Despite the apparent unity of the organism, occasionally alleles can
gain an evolutionary advantage at a cost to  individual fitness
\citep{Burt2006}, often by exploiting meiosis and gametogensis.
%A clear opportunity for this conflict arises during gametogenesis where alternative alleles are in direct competition for representation in the next generation.
Female meiosis, an asymmetric event in which only one of two alternate alleles
enters the  egg, with the others consigned to the polar body, is one such occasion \citep{Sandler1957,Pardo-ManuelDeVillena2001a}. 
%Female meiosis is particularly unfair, only one of the four products of meiosis becomes an egg, while the other threes are disgarded into polar bodies. 
An allele that biases female meiosis in its favor (i.e. a meiotic driver), may increase in frequency even if this driver entails a pleiotropic fitness cost \citep{Prout1973}, generating a genetic conflict between the success of the driver and organismal fitness.
%Alleles that can segregate into the egg more than half the time in heterozygotes, by exploiting asymmetries, 
%can potentially increase in frequency in the population (i.e. experiencing true meiotic drive).
%If such alleles have fitness consequences they are a source of genetic conflict, 
%and can become balanced in the population if their host in homozygotes 
%out weighs their ability to spread to spread through heterozygotes. 
Meiotic drivers observed in nature (in both plants
\citep{Buckler1999,Fishman2005,Fishman2008}, and animals
\citep{Agulnik1990,Wu2005,Pardo-ManuelDeVillena2001c}) highlight this
conflict -- the selfish benefits of drive and the associated
pleiotropic fitness costs sustain a balanced polymorphism
\citep{Prout1973}, 
and often generate on ongoing evolutionary escalation of drive suppressors and enhancers \citep{Dawe1996,Fishman2008}. 
%Such balanced drive systems can cause the subsequent evolution of linked enhancers of drive
%and of supressors of drive throughout the genome. 
%Indeed, the known polymorphic female meiotic drive systems (OM, maize knobs, IN, mimulus, others?) habor
%a great diversity of enhancers and supressors. 
The threat of meiotic drive to organismal fitness is potentially so
severe that it has been hypothesized that many basic properties of
meiosis and o\"{o}genesis, including the initial genome doubling in
meiosis I \citep{Haig1991}, arrested female meiosis
\citep{Mira1998}, centromere mechinery \citep{MalikHenikoff}, and sex differences in the recombination rate \citep{Haig2010,Brandvain2012} have evolved to disrupt meiotic drive and enforce fairness.

@@

It is also not the case that sperm is mechanistically incapable of influencing the outcome of female meiosis.
%This is not just idle speculation, as there is also direct evidence that sperm components and indeed different sperm genotypes can affect female meiosis.
%are mechanistic reasons to think that this is more than idle speculation.
Mechanistic evidence for this possibility comes from
\emph{C. elegans}, where experimentally suppressing XXX leads to
premature deployment of the aster (a vital component of mitotic
machinery) provided by the sperm, disrupting MII meiotic segregation in the egg, leading to a triploid zygote. 
%This demonstrates that physical components derived from the sperm are are capable of affecting the outcome of female meiosis.
Additionally, genetic evidence suggests that the transmission patterns
in heterozygous females many potentially depend on sperm haplotype. 
Specifically, the two best characterized female meiotic drive systems in mouse (In and Om), both operate by distorting the second meiotic division, 
and in both systems the outcome of female meiosis depends the genotype of the fertilizing sperm \citep{Agulnik1993,Wu2005}.

@@

\subsection*{ A) Invasion of the population by a driving allele that promotes
itself.}

The standard model of female meiotic drive is two allele model, where the
allele $1$ is transmitted $d$ fraction of the time through female
meiosis ($d > \frac{1}{2}$), regardless of the genotype of the sperm. 
In mammals, fertilization takes place at MII, so we imagine this drive must be taking place at MII in order
for a sperm to have any influence. For drive to take place at MII there has to be an uneven number of
 crossovers between the centromere and the drive locus, such that realistically $d$ is bounded to be $<0.75$?. 
In other systems where fertilization occurs during MI, sperm could influence either drive at MI or MII, 
and drivers at MI can have a $d=1$ if they occur in tight linkage with the centromere.

@@

When conditions are suitable for such alleles to spread into the
population, they spread slowly through low frequencies as they rarely
drive as they are unlikely to be fertilized by a sperm with the right
allele. Once they become established they spread rapidly to fixation due to their
positive frequency dependent behavior (Inset in Figure \ref{Invasion_space}). 
Unlike standard meiotic drive systems 
these alleles can not become balanced in the population  by homozygote
cost, as they have already over come this cost just to enter the population.

Things are even worse for the sperm-dependent driver if it suffers any
fitness cost in the heterozygote. Our standard driver can invade the
population from low frequency if its drive in heterozygotes out weighs
the fitness cost they suffer (XXX). However, when at low frequency our
sperm dependent driver is often in heterozygotes but will only drive
if fertilized by the same allele. If our allele incurs even very weak costs
in heterozygotes it can be prevented from spreading from very low
frequencies, unless its drive is very strong. The dynamics of 
our sperm-dependent drivers with an additive cost resemble an underdominant polymorphism
(in that the system is bistable). Below a cutoff frequency the allele is
unable to enter the population (in a deterministic model) as it is
paying the cost of the drive but drive is ineffective (see Figure
S\ref{bistable_additive}). Above this
frequency the allele increases deterministically as its drive is sufficiently effective 
becayse heterozygote eggs are fertilized by sperm with our allele
sufficiently often to beat the costs. 
This bistability suggests that sperm-dependent drivers will often be 
unable to enter into reasonably sized populations, and if they do they
should rapidly be spread fixation.

A model where the sperm allele affects outcome of female meiosis may not be biologically
realistic as there is very little sperm expression. More realistically the male genotype that produced the sperm
may be more relevant, as males could place products in their sperm that influenced the outcome of female meoisis.
However, in practice models that allow the influence of fertilizing male genotype on female meiosis 
seem to behave qualitatively to those based on the allele of the
fertilizing sperm (see Supplementary Figures XXX and XXX).

Given the difficulty that meotic drive alleles that promote themselves
through sperm have even entering the population, and the speed at
which they fix if they do, it seems very unlikely that such alleles
could drive the evolution of female

@@

\begin{figure}
\includegraphics[width = 0.8 \textwidth]{Figures/trajectories_of_sperm_based_supressors.eps} 
\caption{Trajectories of two sperm-based supressors of drive. Possibly
merge this figure the 3 allele cartoon}  \label{Trajectories_of_supressors}
\end{figure}

\subsection*{B) A more biologically realistic selfish sperm-egg haplotype system}
These self promoting alleles are somewhat biologically unrealistic as
we are hypothesizing a newly evolved allele  
that both drives in female meiosis and has the ability to influence that drive in sperm.
Perhaps it is more realistic to think that a female drive system first
evolves and is maintained as a polymorphism in the population. 
This can then allow another allele to appears that further facilitates
female meiotic drive through its action in the sperm.

Our sperm-acting allele will be best able to spread if it arises in very close
linkage with the initial meiotic drive system as then it could potentially
hitchhiking due to the additional drive it causes. Polymorphic transmission distortion systems
have often evolved complex sets of inversions to prevent their breakup
\citep{XXX}, so it is reasonable to assume that for some female drive
systems there may be a reasonably large mutational target allowing our
allele to arise in tight linkage. Assuming very tight linkage we can
model this system using a model with three alleles at the same
locus. The first two alleles ($A$ and $B$) consist of our ancestral non-driver and
our original driver allele. The third allele ($C$) is the 
combined haplotype of our driver allele and the fully linked
sperm-acting promotor of drive.
 See Figure
\ref{Eggsperm_3_allele_cartoon} for a simple schematic of this drive model.

The problem for our sperm-acting promoter of drive (the novel C
allele), is that in order to have had time to evolve our original
drive system ($B$) has to be trapped at drive-homozygote disadvantage selection balance. 
Our new allele has arisen on the drive background so it
suffers from the same disadvantage. 
Worse yet, when our new allele it is rare it is often not helping
itself drive in females but instead $B$ ($C$ sperm meeting $AB$ eggs).
As $C$ facilitiates $B$ these interactions will frequently form 
$BC$ heterozygote which suffers from the full
effect fitness costs as $BB$.

Sperm-acting selfing alleles are therefore at a profound disadvantage
in this scenerio, even more so than under the previous two allele model.
We have found no parameter range of this
three allele system that allows the selfish sperm-acting allele $C$ to
invade the population.

There are other types of sperm acting alleles that influence
female meiosis that can invade the population. 
In particular sperm-acting alleles that act to restore 
the fairness of meiosis in females can spread. 
We could imagine that these act by 
disrupting whatever mechanistic asymmetry in female meiotic segregation the
drive initially evolved to exploit \citep{Pardo-ManuelDeVillena2001a}.\\

If these sperm-acting disrupting alleles arise at unlinked loci or on
the $A$ background they can readily invade a population segregating
for the drive system ($AB$, see Figure ). The invasion of such allele
will lower the frequency of the original drive system (perhaps to zero)
and will spread to fixation if they do not carry strong fitness costs
(Figure \ref{Eggsperm_3_allele_cartoon} and \ref{Trajectories_of_supressors}A). \\
%These alleles are the sperm-acting equivalent of female 
%supressor of drive.

Intriguingly these sperm-acting distrupting alleles can also spread
when they arise tightly linked to the original drive system (B), forming
an novel third allele/haplotype C (Figures \ref{Eggsperm_3_allele_cartoon} and \ref{Trajectories_of_supressors}B). This new allele spreads
to displace the original drive system (B) as it benefits from drive
but by disrupting drive when it is in the sperm it forms
 the deleterious $BC$ and $CC$ combinations less often (than the
 analogous combinations for the $B$ allele). 
\gc{Do we mention the In and Om locus here or in the discussion as a
  potential e.g. Do we mention the fact that the frequency of these
  newly invade sperm dodging alleles can be lower than the original system?}\\

Finally given a system of a balanced meiotic drive system which avoid
its own ill effects through a sperm-based mechanism
, such as that depicted in Figure XX and XX, female-based modifiers 
can evolve that act to further facilitate the sperms action in
disrupting driver.

\section*{Discussion}

\section*{OUTLINE}
\yb{YB: To me its (1) male enhancement of female drive cannot maintain a stable polymorphisms, and (2) An allele in sperm can evolve to suppress its drive in females.}
show that such alleles:
\begin{enumerate}
\item can't be balanced, \\
\item and homozygous problems are tested out at low freq.  \\
\item Any heterozygous problems, leads to a bistable allele\\
\item If these alleles take off they speed through to fixation\\
\item If allele has any drive ability in absence of sperm effect that is what allows it to enter the population
and sperm effect isn't a further cause of conflict. What if anything do we mean by this?\\
\item PERHAPS HERE WE INTRODUCE A SELF-RESTRAINING ALLELE
\end{enumerate}

@@
